### 2014-05-18 patchy-selection-paper.tex
edited 2.4; slight figure modifications

Consider a population spread across a landscape, 
within which are patches of habitat to which individuals are (initially) poorly adapted,
surrounded by large areas to which the population is well adapted.
(When we refer to ``patches'' it is to these pieces of poor habitat.)
Suppose it takes only a single mutational change to create an allele
($B$) that adapts an individual to the poor habitat type.
The required change could be as specific as a single base change, 
or as general as a knock out of a gene or one of a set of genes.
We assume that this mutational change occurs at a (low) rate of $\mu$ per chromosome per generation,
and that this initially rare new allele has fitness $1+s_p$ relative to the unmutated type ($b$) in these ``poor'' habitat patches.
Outside of these patches we assume that the new allele has a relative fitness
$1-s_m$  when rare in the intervening areas, with $s_p$ and $s_m$ both positive.
To simplify matters, we assume that the disadvantage $s_m$ 
is sufficiently large that, on the relevant timescale,
the allele is very unlikely to fix locally in the regions where it is at a disadvantage.
(The case where $s_m=0$ requires a different approach, which we do not treat here.)

@@

\gc{I'd suggest making Figure 2 and Fig 1A all part of same figure
  they are showing the same basic thing good fit of equation of
  cline. The s's in 1A should be labeled sm and sb. I'd suggest that
  the second part of 1 become its own figure or be paired with fig.3
  as they are both about rare excursions. Could you make the CI or
  sample path in Figure 1B be different colors, the grey and black are
too close to each other.}
\plr{how's that?  I'm not sure what to do about the grey; the real problem is that the CIs are too small. If I ran them for less time they would be\dots}

@@

\begin{figure}[ht!]
  \begin{center}
    \includegraphics{sim-occupation-freqs}
  \end{center}
  \caption{
  \textbf{(A)} 
  Mean occupation frequencies of individual demes (grey dots)
  and averaged over demes at similar distances (black circles), on
  a one-dimensional grid of 201 demes, and
  a two-dimensional grid of $101\times 101$ demes.
  Superimposed in black is the decay predicted by \eqref{eqn:eqfreq},
  with $C$ chosen arbitrarily to give a reasonable fit.
  \textbf{(B)} 
  Fluctuations: colored lines are snapshots of allele frequencies at evenly spaced times
  from a one-dimensional grid of 151 demes;
  the solid lines shows the cumulative mean occupation frequency across all generations.
  For both plots, the simulation was run for 1000 generations;
  the focal allele is beneficial in the central 20 demes, with $s_p=0.1$,
  and deleterious elsewhere, with $s_m=-0.02$; 
  demes had 1000 individuals in each;
  dispersal was to nearby demes with a mean dispersal distance of $\sigma \approx 1.66$  times the deme spacing.
  }   \label{fig:sim_occupation_freqs}
\end{figure}

@@

To understand the process of adaptation by migration,
we therefore need to look more closely at the rare families of $B$ alleles far from the patch where it is established.
Consider one such family,
that lives far from the original patch.
As noted above, we will model this by assuming that each individual
migrates, then reproduces, independently of all other family members.
The mean squared migration distance is $\sigma^2$, takes the migrant in a random direction,
and if migration takes the individual into a patch where the $B$ allele is favored, 
we drop the individual from our family.
Since the $B$ allele is uniformly deleterious elsewhere,
we can decouple reproduction and migration
by first forming a tree in which each individual has a random number of offspring
(with mean $1-s_m$ and variance $\xi^2$),
then recursively assigning spatial locations to each individual in the tree,
and finally pruning any branches that happen to wander into a patch.
This pruning removes both branches leading back into the already established patch (which are irrelevant)
and branches leading into the new patch (we will need to count these).
Since we are concerned with the rare migrant families that transit quickly between patches, 
back migrants to the already established patch will contribute little, since they are going the wrong direction;
so to simplify the analysis we ignore this detail.  % but could add this detail in an appendix

Since each individual has on average $(1-s_m)$ offspring,
the expected family size after time $t$ is $(1-s_m)^t$.
Therefore, if we define $k_e(t)$ to be the chance that the family
has gone extinct by time $t$, 
\plr{switched from $p_e(t)$ to $k_e(t)$ to not conflict with prob of estab.  better suggestion?}
and let $K_t$ be the (random) family size conditioned on nonextinction,
then by Bayes' rule, $(1-s_m)^t = (1-k_e(t))\E[K_t]$.
It turns out that if the distribution of offspring numbers is not too heavy-tailed \citep[see][for details]{jagers1975branching},
then $K_t$ has a limiting distribution: $K_t \dconv K$ 
and that $(1-s_m)^{t}/(1-k_e(t)) \to \E[K]$ as $t \to \infty$.
In other words, the chance of survival to time $t$ is a constant multiplied by $(1-s_m)^t$,
and conditional on survival, the family size is given by a fixed distribution.

@@

First consider equation \eqref{eqn:gestalt_q}.
Suppose that $Z$ is one such spatial branching process as above, started at time 0 with a single individual at $x$,
defined so that $Z_t(A)$ is the number of descendant individuals alive at time $t$ living in the spatial region $A$.
The \emph{mean occupation density} of $Z$ can be thought of 
as the expected total number of offspring of a family beginning at $x$ that ever live at $y$.
Concretely, it is the density of $\E[\int_0^t Z_t dt]$,
i.e.\ the smooth function $u(x,y)$ that, integrated over a set $A$, 
gives the expected total number of individuals ever living in $A$:
\begin{align} \label{eqn:occupation_density_defn}
    \E\left[ \int_0^\infty Z_t(A) dt \right] = \int_A u(x,y) dy .
\end{align}
\gc{do we need to talk about u a little more it }
\plr{not sure what to say}
Since reproduction and spatial motion are independent,
$u(x,y)$ is the sum across all generations of the expected number of individuals alive in that generation,
multiplied by the probability density that spatial motion takes a lineage from $x$ to $y$,
i.e.\
\begin{align}
    u(x,y) = \int_0^t (1-s_m)^t p_t(x,y) dt .
\end{align}
The equilibrium frequency as decomposed in expression \eqref{eqn:gestalt_q} 
is the sum of mean occupation densities of a constant outflux of branching processes
from $\partial B(r_0)$:
\begin{align} \label{eqn:occupation_integral}
    q(y) &= \int_{\partial B(r_0)} \gamma(x) u(x,y) dx  . % \\
         % &= \int_{\partial B(r_0)} \gamma(x) \int_0^t (1-s_m)^t p_t(x,y) dt dx .
\end{align}
\gc{define $\gamma(x)$ presumably the outflux of branches leaving ball}
\plr{it was defined at the start of this paragraph.}

@@

Intuitively, $h(x,y)$ counts each family once if it hits the patch,
while the occupation density $u(x,y)$ counts the total amount of time that the family spends in the patch.
Since the expected total time spent in the patch does not depend on the distance at which the family started
if we know that it survived to hit the patch,
it is intuitively reasonable that $u(x,y)$ and $h(x,y)$ are simply related.
Following this intuition,
and that $(1-k_e(t)) \approx (1-s_m)^t/\E[K]$,
it turns out (see Appendix~\ref{apx:hitting_occupation}) that
\begin{align} 
  h(x,y) = \frac{ g(A) }{ \E[K] } u(x,y),
\end{align}
where $g(A)$ is the total expected occupation time of the patch given the family arrives,
and recall that $\E[K]$ is the expected family size after a large number of generations, given survival.
Since the integrands in expressions \ref{eqn:occupation_integral} and \ref{eqn:migrate_integral} only differ by a constant factor
(at least asymptotically as $x$ and $y$ are far apart),
we now know that the outflux of successful migrant families is equal to the mean occupation frequency
multiplied by $g(A) f(A) / \E[K]$.
The probability $f(A)$ that the family does not die out once it arrives at the new patch is
$\E[1-(1-p_e)^K] \approx 2 s_p \E[K] / \xi^2$,
if each member arriving has chance $p_e$ of establishing, independently.
% If we let $w$ be the width of the patch, then the remaining term is
% \begin{align}
%   s(w) \simeq \frac{1}{2\alpha} \left( 1 - \left( \frac{\pi w \sqrt{2 \alpha}}{ 2 \sigma } \right)^{\frac{d-1}{2}} e^{-w \sqrt{2\alpha}/\sigma} \right).
% \end{align}
Combining these terms with the expression for $g(A)$ from Appendix~\ref{apx:hitting_occupation},
\begin{align}
  \label{eqn:migrate}
  \migrate(y) \approx q(y) \times \frac{ \rho s_p  }{ \xi^2 \log(1/(1-s_m)) } 
  % \left( 1 - \left( \frac{\pi w \sqrt{2 \alpha}}{ 2 \sigma } \right)^{\frac{d-1}{2}} e^{-w \sqrt{2\alpha}/\sigma} \right).
\end{align}
For patches that are small relative to $\sigma/\sqrt{2s_m}$, there is an additional correction term (see Appendix~\ref{apx:hitting_occupation}),
but small patches violate other assumptions above, so we omit it.

@@

We will assume that once a $B$ allele is introduced (by migration or mutation) and successfully begins to establish,
it becomes established in the poor habitat patch rapidly. 
Under this approximation, and the ``selective exclusion'' just discussed,
after one of the patches becomes colonized by a single $B$ allele, 
the other patch will become colonized by migration or mutation, but not by both. 
As such, the question of how the the second patch adapts
simply comes down to whether a new mutation or a migrant allele is the first to become established in the second patch. 
Taking expressions \eqref{eqn:mutrate} and \eqref{eqn:migrate} above,
we see that if the patches each have sufficiently large area,
the rate of migration and mutation will be on the same order if 
$A \mu \approx (1/s_m) \exp(- R \sqrt{2 s_m} / \sigma)$,
where $R$ is the distance between the patches,
and recall that $A$ is the area of the not-yet-adapted patch.
In other words, the migrational analogue of ``mutational influx'' ($\mu \rho A$) 
is $\rho (1/s_m) \exp(- R \sqrt{2 s_m} / \sigma)$,
which depends fairly strongly on the selective disadvantage $s_m$ between patches.
Equivalently, migration will occur much quicker than mutation 
if $R/\sigma > (\log(1/s_m) - \log(A\mu))/\sqrt{2s_m}$.
\gc{the two paragraphs either side of this will need some reworking.}
\plr{want to do that? or be a bit more specific? what works/doesn't?}

If we take the mutational influx in a patch to be one per 1,000
generations ($A \mu = .001$), \gc{don't think this is mutational
  influx as it doesn't have $\rho in it$}
and the selection against transiting alleles $s_m = .01$,
adaptation procedes independently for distances $R/\sigma > \log(s_m)/\sqrt{s_m}$,
so if the selective pressure is fairly strong (say, $s_m=.05$),
then the distance between patches must be about 30 times the dispersal distance (i.e.\ $R/\sigma > 30$),
while if $s_m$ is much smaller, say $s_m = .001$, 
then the patches must be separated by several hundred times the dispersal distance (i.e.\ $R/\sigma > 300$).
The situation does not change much if the influx is quite a bit higher -- even if $A \mu = 1$,
the required separation between patches is only reduced by $7/\sqrt{s_m}$,
to $R/\sigma > 10$ with $s_m=.05$ and $R/\sigma>150$ for $s_m=.001$.
If the separation between patches is much larger than this transitional value -- certainly, by a factor of ten -- 
then patches are effectively isolated -- most adaptation will occur independent between patches.
On the other hand, if the separation is much smaller, then a single
mutation is likely to fix in both patches.

One way to visualize this is by a phase diagram.
In figure \ref{fig:phase_diagram},
we give a contour plot for the time until adaptation by migration ($\Tmig = 1/\migrate$) and by new mutation ($\Tmut=1/\mutrate$)
as a function of various parameters. \gc{Parallel mutation will occur
where $\Tmut \gg \Tmig $ (reder areas), while adpatation by migration will occur
where $\Tmig \gg \Tmut $ (bluer areas).}

\begin{figure}[ht]
  \begin{center}
    \includegraphics{phase-diagram-log}
  \end{center}
  \caption{
  Shaded contour plots of
  mean time until adaptation in a new patch occurs by migration ($\Tmig$, blue) or by mutation ($\Tmut$, red),
  as a function of population density ($\rho$), 
  distance to an already adapted patch in units of dispersal distance ($R/\sigma$),
  strength of deleterious selection ($s_m$),
  and mutation rate scaled by patch area ($\mu A$).
  The line at which $\Tmig=\Tmut$ is shown in black.
  Default parameter values were $A=100$, $\rho=10$, $\mu=10^{-8}$, $s_p=0.1$, $s_m=0.01$, $R/\sigma=50$, and $\xi^2=1$.
  Note that population density and mutation rate are shown on a log scale.
  \label{fig:phase_diagram}
  }
\end{figure}

If we are willing to take these approximations at face value, 
and assuming both rates are small, 
the time until the first of the two patches is colonized by the $B$
allele will be approximately exponentially distributed with mean $1/(2
\mutrate)$.
Following this, the time till the second patch is subsequently colonized 
(via either migration or new mutation) 
will be exponentially distributed with mean $1/(\mutrate+\migrate)$.
Both these times scale linearly with $\xi^2/(\rho s_{B})$, 
so that the overall time to adaptation is robustly predicted to increase with offspring variance ($\xi^2$)
and decrease with population density and benefical selection coefficient.
Furthermore, the probability that the second adaptation is a new mutation,
i.e.\ the probability of parallel adapatation, is in $d=1$ ($d=2$ only differs slightly)
\begin{equation}
  \frac{\mutrate}{\mutrate+\migrate(r)} = \frac{ A \mu \alpha }{2 A \mu \alpha +
    \left(r \sqrt{2 s_m} r / \sigma \right)^{(d-1)/2} \exp\left(- \sqrt{2 s_m} r / \sigma \right) },
\end{equation}
so that the probability of parallel mutation should increase
approximately logistically with the distance between the patches, at rate $\sqrt{s_m} /\sigma$.

>>>>

Consider a population spread across a landscape, 
within which are patches of habitat to which individuals are (initially) poorly adapted,
surrounded by large areas to which the population is well adapted.
(When we refer to ``patches'' it is to these pieces of poor habitat.)
Suppose it takes only a single mutational change to create an allele
($B$) that adapts an individual to the poor habitat type.
The required change could be as specific as a single base change, 
or as general as a knock out of a gene or one of a set of genes.
We assume that this mutational change occurs at a (low) rate of $\mu$ per chromosome per generation,
and that this initially rare new allele has fitness $1+s_p$ relative to the unmutated type ($b$) in these ``poor'' habitat patches.
Outside of these patches we assume that the new allele has a relative fitness
$1-s_m$  when rare in the intervening areas, with $s_p$ and $s_m$ both positive.
(Here ``fitness'' we take to be the intrinsic growth rate of the type when rare.)
To simplify matters, we assume that the disadvantage $s_m$ 
is sufficiently large that, on the relevant timescale,
the allele is very unlikely to fix locally in the regions where it is at a disadvantage.
(The case where $s_m=0$ requires a different approach, which we do not treat here.)

@@

\gc{Could you make the CI or sample path in Figure 1B be different colors, the grey and black are too close to each other.}
\plr{how's that?  I'm not sure what to do about the grey; the real problem is that the CIs are too small. If I ran them for less time they would be\dots}

@@

\begin{figure}[ht!]
  \begin{center}
    \includegraphics{sim-occupation-freqs}
  \end{center}
  \caption{
  \textbf{(A)} 
  Mean occupation frequencies of individual demes (grey dots)
  and averaged over demes at similar distances (black circles), on
  a one-dimensional grid of 201 demes, and
  a two-dimensional grid of $101\times 101$ demes.
  Superimposed in black is the decay predicted by \eqref{eqn:eqfreq},
  with $C$ chosen arbitrarily to give a reasonable fit.
  (The uptick at the right end is due to boundary effects near the edge of the range.)
  \textbf{(B)} 
  Fluctuations: colored lines are snapshots of allele frequencies at evenly spaced times
  from a one-dimensional grid of 151 demes;
  the solid lines shows the cumulative mean occupation frequency across all generations.
  For both plots, the simulation was run for 1000 generations;
  the focal allele is beneficial in the central 20 demes, with $s_p=0.1$,
  and deleterious elsewhere, with $s_m=-0.02$; 
  demes had 1000 individuals in each;
  dispersal was to nearby demes with a mean dispersal distance of $\sigma \approx 1.66$  times the deme spacing.
  }   \label{fig:sim_occupation_freqs}
\end{figure}

@@

To understand the process of adaptation by migration,
we therefore need to look more closely at the rare families of $B$ alleles far from the patch where it is established.
Consider one such family,
that lives far from the original patch.
As noted above, we will model this by assuming that each individual
migrates, then reproduces, independently of all other family members.
The mean squared migration distance is $\sigma^2$, takes the migrant in a random direction,
and if migration takes the individual into a patch where the $B$ allele is favored, 
we drop the individual from our family.
Since the $B$ allele is uniformly deleterious elsewhere,
we can decouple reproduction and migration
by first forming a tree in which each individual has a random number of offspring
then recursively assigning spatial locations to each individual in the tree,
and finally pruning any branches that happen to wander into a patch.
By our definition of fitness, the mean number of offspring per individual is $\exp(-s_m)\approx 1-s_m$,
and the variance is $\xi^2$.
\plr{I changed this since equation \eqref{eqn:eqfreq} is in the small-$s_m$ limit, 
and therefore doesn't differentiate between $s_m$ and $\log(1/(1-s_m))$;
doing it this way means we don't have to define another parameter.}
This pruning removes both branches leading back into the already established patch (which are irrelevant)
and branches leading into the new patch (we will need to count these).
Since we are concerned with the rare migrant families that transit quickly between patches, 
back migrants to the already established patch will contribute little, since they are going the wrong direction;
so to simplify the analysis we ignore this detail.  % but could add this detail in an appendix

Since each individual has on average $e^{-s_m}$ offspring,
the expected family size after time $t$ is $e^{t s_m}$.
Therefore, if we define $k_e(t)$ to be the chance that the family
has gone extinct by time $t$, 
\plr{switched from $p_e(t)$ to $k_e(t)$ to not conflict with prob of estab.  better suggestion?}
and let $K_t$ be the (random) family size conditioned on nonextinction,
then by Bayes' rule, $e^{-s_m t} = (1-k_e(t))\E[K_t]$.
It turns out that if the distribution of offspring numbers is not too heavy-tailed \citep[see][for details]{jagers1975branching},
then $K_t$ has a limiting distribution: $K_t \dconv K$ 
and that $e^{-s_m t}/(1-k_e(t)) \to \E[K]$ as $t \to \infty$.
In other words, the chance of survival to time $t$ is a constant multiplied by $e^{-s_m t}$,
and conditional on survival, the family size is given by a fixed distribution.

@@

First consider equation \eqref{eqn:gestalt_q}.
Suppose that $Z$ is one such spatial branching process as above, started at time 0 with a single individual at $x$,
defined so that $Z_t(A)$ is the number of descendant individuals alive at time $t$ living in the spatial region $A$.
The \emph{mean occupation density} of $Z$ can be thought of 
as the expected total number of offspring of a family beginning at $x$ that ever live at $y$.
Concretely, it is the density of $\E[\int_0^t Z_t dt]$,
i.e.\ the smooth function $u(x,y)$ that, integrated over a set $A$, 
gives the expected total number of individuals ever living in $A$:
\begin{align} \label{eqn:occupation_density_defn}
    \E\left[ \int_0^\infty Z_t(A) dt \right] = \int_A u(x,y) dy .
\end{align}
\gc{do we need to talk about u a little more it }
\plr{not sure what to say}
Since reproduction and spatial motion are independent,
$u(x,y)$ is the sum across all generations of the expected number of individuals alive in that generation,
multiplied by the probability density that spatial motion takes a lineage from $x$ to $y$,
i.e.\
\begin{align}
    u(x,y) = \int_0^t e^{-s_m t} p_t(x,y) dt .
\end{align}
The equilibrium frequency as decomposed in expression \eqref{eqn:gestalt_q} 
is the sum of mean occupation densities of a constant outflux of branching processes
from $\partial B(r_0)$:
\begin{align} \label{eqn:occupation_integral}
    q(y) &= \int_{\partial B(r_0)} \gamma(x) u(x,y) dx  . % \\
         % &= \int_{\partial B(r_0)} \gamma(x) \int_0^t e^{-s_m t} p_t(x,y) dt dx .
\end{align}
\gc{define $\gamma(x)$ presumably the outflux of branches leaving ball}
\plr{it was defined at the start of this paragraph.}

@@

Intuitively, $h(x,y)$ counts each family once if it hits the patch,
while the occupation density $u(x,y)$ counts the total amount of time that the family spends in the patch.
Since the expected total time spent in the patch does not depend on the distance at which the family started
if we know that it survived to hit the patch,
it is intuitively reasonable that $u(x,y)$ and $h(x,y)$ are simply related.
Following this intuition,
and that $(1-k_e(t)) \approx e^{-s_m t}/\E[K]$,
it turns out (see Appendix~\ref{apx:hitting_occupation}) that
\begin{align} 
  h(x,y) = \frac{ g(A) }{ \E[K] } u(x,y),
\end{align}
where $g(A)$ is the total expected occupation time of the patch given the family arrives,
and recall that $\E[K]$ is the expected family size after a large number of generations, given survival.
Since the integrands in expressions \ref{eqn:occupation_integral} and \ref{eqn:migrate_integral} only differ by a constant factor
(at least asymptotically as $x$ and $y$ are far apart),
we now know that the outflux of successful migrant families is equal to the mean occupation frequency
multiplied by $g(A) f(A) / \E[K]$.
The probability $f(A)$ that the family does not die out once it arrives at the new patch is
$\E[1-(1-p_e)^K] \approx 2 s_p \E[K] / \xi^2$,
if each member arriving has chance $p_e$ of establishing, independently.
% If we let $w$ be the width of the patch, then the remaining term is
% \begin{align}
%   s(w) \simeq \frac{1}{2\alpha} \left( 1 - \left( \frac{\pi w \sqrt{2 \alpha}}{ 2 \sigma } \right)^{\frac{d-1}{2}} e^{-w \sqrt{2\alpha}/\sigma} \right).
% \end{align}
Combining the terms with expressions for $g(A)$ in Appendix~\ref{apx:hitting_occupation},
we obtain that
\begin{align}
  \label{eqn:migrate}
  \migrate(y) &\approx q(y) \times \frac{ \rho s_p  }{ \xi^2 s_m }  \\
  % \left( 1 - \left( \frac{\pi w \sqrt{2 \alpha}}{ 2 \sigma } \right)^{\frac{d-1}{2}} e^{-w \sqrt{2\alpha}/\sigma} \right).
  &\approx C \frac{ \rho s_p  }{ \xi^2 s_m } \left( |x| \sqrt{2 s_m}/\sigma \right)^{-(d-1)/2} \exp( - |x| \sqrt{2 s_m} / \sigma) .
\end{align}
This does not depend on $A$, because although
$f(A)$ and $g(A)$ in principle depend on the patch size (and geometry),
the dependence is very weak:
for instance, the width of a circular patch only very weakly affects the chance 
of establishment of a new migrant that appears on its edge,
as long as the patch is large enough.
For more discussion, and correction terms, see Appendix~\ref{apx:hitting_occupation}.

@@

We will assume that once a $B$ allele is introduced (by migration or mutation) and successfully begins to establish,
it becomes established in the poor habitat patch rapidly. 
Under this approximation, and the ``selective exclusion'' just discussed,
after one of the patches becomes colonized by a single $B$ allele, 
the other patch will become colonized by migration or mutation, but not by both. 
As such, the question of how the the second patch adapts
simply comes down to whether a new mutation or a migrant allele is the first to become established in the second patch. 
To work out how adaptation is likely to proceed, 
we can compare expressions \eqref{eqn:mutrate} and \eqref{eqn:migrate} above
for the effective rates of adaptation by new mutation and by migration.

The first thing to notice is that effective migration and mutation rates will be on the same order if 
$A \mu \approx (1/s_m) \exp(- R \sqrt{2 s_m} / \sigma)$,
where recall that $R$ is the distance between the patches,
and $A$ is the area of the not-yet-adapted patch.
In other words, the migrational analogue of ``mutational influx'' ($\mu \rho A$) 
is $\rho (1/s_m) \exp(- R \sqrt{2 s_m} / \sigma)$,
which depends fairly strongly on the selective disadvantage $s_m$ between patches.
Equivalently, the rates are roughly equal if
$R/\sigma = \log(1/(s_m A \mu))/\sqrt{2s_m}$,
which gives the critical gap size past which adaptation will be mostly parallel
in terms of selection coefficients, patch size, and mutation rate.

% ss <- 10^(-(1:6))
% log(1/ss)
% 1/sqrt(2*ss)
% sm <- .05; (-log(c(1e-3,1))-log(sm))/sqrt(2*sm)
% sm <- .001; (-log(c(1e-3,1))-log(sm))/sqrt(2*sm)
% (7+log(1/ss))/sqrt(2*(ss))
% (7+log(1/(.5*ss)))/sqrt(2*(.5*ss))
If we take the mutational influx in a patch to be one per 1,000 generations ($\rho A \mu = .001$)
and the population density to be 1 ($\rho =1$, so $\log(A\mu)\approx 7$)
then adaptation is mostly parallel for patches separated by gaps larger than $R/\sigma > (7+\log(s_m))/\sqrt{s_m}$.
If the selective pressure is fairly strong (say, $s_m=.05$),
then for migration to be slower than mutation,
the distance between patches must be at least 30 times the dispersal distance (i.e.\ $R/\sigma > 30$).
If $s_m$ is much smaller, say $s_m = .001$, 
then the patches must be separated by several hundred times the dispersal distance (i.e.\ $R/\sigma > 300$).
The situation does not change much if the influx is quite a bit higher -- even if $A \mu = 1$ (so $\log(A\mu) \approx 2.3$),
the required separation between patches is only reduced to $R/\sigma > 10$ with $s_m=.05$ and $R/\sigma>150$ for $s_m=.001$.
On the other hand, if the separation is much smaller than this transitional value, 
then adaptations are likely to be shared between patches.

One way to visualize this is by a phase diagram.
In figure \ref{fig:phase_diagram},
we give a contour plot for the time until adaptation by migration ($\Tmig = 1/\migrate$) and by new mutation ($\Tmut=1/\mutrate$)
as a function of various parameters. 
This gives us an idea of not only the parameter space where parallel adaptation is likely ($\Tmig \ll \Tmut$),
but also where adaptation by any means takes a very long time ($\Tmig$ and $\Tmut$ both very large),
and important check in applications.

\begin{figure}[ht]
  \begin{center}
    \includegraphics{phase-diagram-log}
  \end{center}
  \caption{
  Shaded contour plots of
  mean time until adaptation in a new patch occurs by migration ($\Tmig$, blue) or by mutation ($\Tmut$, red),
  as a function of population density ($\rho$), 
  distance to an already adapted patch in units of dispersal distance ($R/\sigma$),
  strength of deleterious selection ($s_m$),
  and mutation rate scaled by patch area ($\mu A$).
  The black, dotted line shows values where $\Tmig=\Tmut$.
  Default parameter values were $A=100$, $\rho=10$, $\mu=10^{-8}$, $s_p=0.1$, $s_m=0.01$, $R/\sigma=50$, and $\xi^2=1$.
  Note that population density and mutation rate are shown on a log scale.
  \label{fig:phase_diagram}
  }
\end{figure}

Taking these approximations at face value, 
and assuming both rates are small, 
the time until the first of the two patches is colonized by the $B$
allele will be approximately exponentially distributed with mean $1/(2
\mutrate)$.
Following this, the time until the second patch is subsequently colonized 
(via either migration or new mutation) 
will be exponentially distributed with mean $1/(\mutrate+\migrate)$.
Both these rates scale linearly with population density ($\rho$) 
and the probability of establishment of a rare allele ($p_e\approx 2 s_b/\xi^2$),
so that the overall time to adaptation is robustly predicted to increase with offspring variance ($\xi^2$)
and decrease with population density and benefical selection coefficient.
Furthermore, the probability that the second adaptation is a new mutation,
i.e.\ the probability of parallel adapatation, is in $d=1$ ($d=2$ only differs slightly)
\begin{equation}
  \frac{\mutrate}{\mutrate+\migrate(r)} = \frac{ A \mu \alpha }{2 A \mu \alpha + \exp\left(- \sqrt{2 s_m} r / \sigma \right) },
\end{equation}
so that the probability of parallel mutation should increase
approximately logistically with the distance between the patches, at rate $\sqrt{s_m} /\sigma$.
